---
title: "Drug Screening Signature Generation"
author: "Sara Gosline"
date: "10/11/2019"
output:
  pdf_document: default
  html_document: default
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(reticulate)
synapse <- import("synapseclient")
syn <- synapse$Synapse()
syn$login()
require(tidyverse)
require(pheatmap)
library(DT)
```

## Drug sensitivity data
The first we want to collect the drug sensitivity data from the NF Data Portal. We collect all the data from digoxin screens, including those from the pNF data.

This analysis tries to build a gene signature of digoxin response using GSVA signatures instead of gene data.


```{r drug data}
#tab<-syn$tableQuery("SELECT distinct model_name,response_type,response,response_unit,symptom_name,disease_name FROM syn20556439 where DT_explorer_internal_id= 11218 and disease_name in ('NF1','no disease') and organism_name='human'")$asDataFrame()
tab<-syn$tableQuery("SELECT distinct model_name,response_type,response,response_unit,symptom_name,disease_name,DT_explorer_internal_id FROM syn20556439 where disease_name in ('NF1','no disease') and organism_name='human'")$asDataFrame()
#datatable(tab)


##lets get drugs that are tested in both NF1 and knockout and have more than one sample
##of each
drugs<-tab%>%subset(response_type!='IC50_abs')%>%
  group_by(response_type,DT_explorer_internal_id)%>%
  mutate(numdis=n_distinct(disease_name))%>%
  mutate(numSamps=n_distinct(model_name))%>%
  subset(numdis==2)%>%
  subset(numSamps>4)

#now look for specific drugs that have differential response in NF1 disease!
dtab<-drugs%>%tibble::rowid_to_column()%>%
  mutate(nresponse=as.numeric(response))%>%
  spread(key=disease_name,value=nresponse)%>%
  group_by(response_type,DT_explorer_internal_id)%>%
  mutate(pval=wilcox.test(`no disease`,NF1)$p.value)%>%
  ungroup()%>%group_by(response_type)%>%
  mutate(correctedP=p.adjust(pval))%>%ungroup()


##now we have some that are significant (not passing testing)
sig.drugs<-subset(dtab,pval<0.005)%>%
  subset(response_type%in%c('Min_viability','AUC_Simpson'))%>%
  dplyr::select(DT_explorer_internal_id,response_type,pval)%>%distinct()

dmap<-syn$tableQuery(paste0("SELECT distinct DT_explorer_internal_id,name from syn18506947 where DT_explorer_internal_id in ('",paste0(unique(sig.drugs$DT_explorer_internal_id),collapse="','"),"')"))$asDataFrame()

pvals.with.names<-sig.drugs%>%left_join(dmap,by='DT_explorer_internal_id')
datatable(pvals.with.names)

```

It's important to remember that the 'no symptom' data includes both NF1 knockout cell lines and WT. So we can compare the NF1 disease vs. non-NF1 cells. But this isn't a great comparison since we have the pNFS included. Here are the results without the pNF data

```{r}
auc.data<- subset(tab,DT_explorer_internal_id%in%unique(sig.drugs$DT_explorer_internal_id))%>%mutate(Response=as.numeric(response))%>%mutate(drugId=as.character(DT_explorer_internal_id))

p<-ggplot(subset(auc.data,response_type%in%c("AUC_Simpson","Min_viability")))+geom_boxplot(aes(x=drugId,y=Response,color=disease_name))+facet_grid(response_type~.)
print(p)
```

Now let's dive into individual compounds so we can explore further. maybe withs ome of their common names
```{r}
for(comp in unique(auc.data$drugId)){
  ddata<-subset(auc.data,response_type%in%c('AUC_Simpson','Min_viability'))%>%
    subset(drugId=comp)
  
  dname=dmap$name[which(dmap$DT_explorer_internal_id==comp)[1]]

  p<-ggplot(ddata)+geom_boxplot(aes(x=response_type,y=Response,color=disease_name))+ggtitle(paste(dname,'(',comp,')response in cells'))
 print(p)
}
```

## Gene Expression Data

Now we can get the gene expression data, map to pathways and find pathways that are unique to specific drug responses. 
```{r gene expression data, echo=FALSE}

query=paste0("SELECT * from syn20370978 where (individualID in ('",
  paste(unique(auc.data$model_name),collapse="','"),
    "') OR specimenID in ('",
  paste(unique(tab$model_name),collapse="','"),"')) ",
  "AND (experimentalCondition is NULL OR experimentalCondition='DMSO' OR experimentalCondition='untreated' or experimentalCondition='')")

gene.tab<-syn$tableQuery(query)$asDataFrame()
gene.tab$individualID[is.na(gene.tab$individualID)]<-gene.tab$specimenID[is.na(gene.tab$individualID)]
gene.tab$individualID<-unlist(gene.tab$individualID)
auc.data<-auc.data%>%rename(individualID='model_name')
#auc.data<-auc.data%>%rename(model_name='individualID')
gene.with.drug<-gene.tab%>%left_join(auc.data,by='individualID')

```

Ok, we have some samples, and now need to compute GSVA on those samples
```{r GSVA, warning=FALSE,echo=FALSE}
require(GSVA)
require(GSVAdata)
data('c2BroadSets')
  library(org.Hs.eg.db)

 
#samps<-dplyr::select(gene.with.drug,disease_name,specimenID,individualID,studyName,response,response_type)%>%
#    distinct()%>%
#  subset(response_type=='AUC_Simpson')
#  samps$studyName<-gsub(" Preclinical Models (Minnesota CCHMC Recombinetics)","",samps$studyName,fixed=T)

#  rownames(samps)<-samps$specimenID
  mat<-reshape2::acast(gene.with.drug,Symbol~individualID,value.var='zScore',fun.aggregate=mean)
  
   map<-AnnotationDbi::select(org.Hs.eg.db,columns=c("SYMBOL",'ENTREZID'),keys=keys(org.Hs.eg.db,'ENTREZID'),multiVals=unique(gene.with.drug$Symbol))
  
  entrez<-map$ENTREZID[match(rownames(mat),map$SYMBOL)]

  
    mat<-mat[which(!is.na(entrez)),]
  rownames(mat)<-entrez[!is.na(entrez)]
  res=gsva(mat,method='ssgsea',gset.idx.list=c2BroadSets)

  dres<-as.data.frame(res)%>%tibble::rownames_to_column('pathway')%>%
      gather('individualID','GSVAscore',-pathway)%>%left_join(auc.data,by='individualID')

  corvals<-dres%>%subset(DT_explorer_internal_id%in%sig.drugs$DT_explorer_internal_id)%>%
    group_by(pathway,DT_explorer_internal_id,response_type)%>%mutate(corVal=cor(GSVAscore,as.numeric(response),method='spearman'))%>%
    mutate(gsvaVar=var(GSVAscore))%>%ungroup()%>%
    dplyr::select(pathway,DT_explorer_internal_id,corVal,response_type,gsvaVar)%>%distinct()
      
  filtered.corvals<-corvals%>%
      subset(abs(corVal)>0.9)%>%#highly correlated
      subset(gsvaVar>0.01)%>%##arbitrary
      subset(response_type=='AUC_Simpson')##let's just stick with this
  
  
```
So we have a lot of data, and i'm not sure what to do with it. For now, we can plot a single drug and the pathways it correlates with. First we plot the AUC. 

```{r plot single drug}
#drugid='277870'
for(drugid in unique(filtered.corvals$DT_explorer_internal_id)){
  filtered.corvals$drugname=drugnames=dmap$name[match(filtered.corvals$DT_explorer_internal_id,dmap$DT_explorer_internal_id)]
  
  drugname=dmap$name[match(drugid,dmap$DT_explorer_internal_id)]

    pathways<- dplyr::select(subset(filtered.corvals,DT_explorer_internal_id==drugid),pathway,corVal)%>%
      arrange(desc(abs(corVal)))%>%
        dplyr::select(pathway)%>%
        unlist()
    
    if(length(pathways)>10)
      pathways=pathways[1:10]


  plot.tab<-dres%>%subset(DT_explorer_internal_id==drugid)%>%
    subset(pathway%in%pathways)%>%
    subset(response_type=='AUC_Simpson')%>%
    mutate('Area Under Curve'=as.numeric(response))%>%
    rename(GSVAscore='GSEA Score')


p<-ggplot(plot.tab)+geom_point(aes(x=`Area Under Curve`,y=`GSEA Score`,col=pathway,shape=disease_name))+ggtitle(paste('Pathways correlated with',drugname,'treatment'))
print(p)
}
```

Because the minimum viability boxplots up top are a little more believable, i will plot those as well.

```{r plot Viability}
#drugid='277870'
for(drugid in unique(filtered.corvals$DT_explorer_internal_id)){

    filtered.corvals$drugname=drugnames=dmap$name[match(filtered.corvals$DT_explorer_internal_id,dmap$DT_explorer_internal_id)]

    drugname=dmap$name[match(drugid,dmap$DT_explorer_internal_id)]

     pathways<- dplyr::select(subset(filtered.corvals,DT_explorer_internal_id==drugid),pathway,corVal)%>%
      arrange(desc(abs(corVal)))%>%
        dplyr::select(pathway)%>%
        unlist()   
    if(length(pathways)>10)
      pathways=pathways[1:10]

  plot.tab<-dres%>%subset(DT_explorer_internal_id==drugid)%>%
    subset(pathway%in%pathways)%>%
    subset(response_type=='Min_viability')%>%
        mutate(`Minimum Viability (%)`=as.numeric(response))%>%
  rename(GSVAscore='GSEA Score')

  p<-ggplot(plot.tab)+geom_point(aes(x=`Minimum Viability (%)`,y=`GSEA Score`,col=pathway,shape=disease_name))+ggtitle(paste('Pathways correlated with',drugname,'treatment'))
  print(p)
}
```
